---
title: "cincinnati syndromic"
author: "Dan Weinberger"
date: "3/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSocrata)
library(lubridate)
library(RCurl)
library(MMWRweek)
library(cdcfluview)
library(gtrendsR)
library(shiny)
library(pbapply)
library(zoo)
library(MASS)
library(tidyr)
library(reshape2)
```

```{r}
script <- getURL("https://raw.githubusercontent.com/weinbergerlab/ExcessILI/master/R/functions.R")
eval(parse(text = script))

script2 <- getURL("https://raw.githubusercontent.com/weinbergerlab/ExcessILI/master/R/aux_functions.R")
eval(parse(text = script2))
```

```{r}
url.cin <- 
  "https://data.cincinnati-oh.gov/resource/vnsz-a3wp.json"
cin1 <- 
  read.socrata(url.cin)

#url.nyc <- "https://data.cityofnewyork.us/resource/76xm-jjuj.json"
#nyc1 <-
#  read.socrata(url.nyc)

# url.marin <- "https://data.marincounty.org/resource/swth-izpe.json"
# marin1 <- read.socrata(
#   url.marin,
#   app_token = "dj52egvrcnreyp83tl797kpqc",
#   email     = "daniel.weinberger@yale.edu",
#   password  = "Yale1234!"
# )

```

```{r, echo=T}
cin1$datetime <- 
  as.POSIXlt(cin1$create_time_incident, #variable with the datetime
                         format=c("%m/%d/%Y %H:%M")) 
cin1$date <-
  date(cin1$datetime)


#These are in protolcol 36: pandemic/epidemic
cin1$flu <-  0
cin1$flu[grep('FLU',cin1$incident_type_desc)] <- 1

#https://wiki.radioreference.com/index.php/Priority_Dispatch_Codes
# 6E1, 6D1-5 6C1-2; 36=flu
cin1$breath2 <- 0
cin1$breath2[grep('BREATHING PROBLEMS',cin1$cfd_incident_type_group)] <- 1
#Mostly fall into 6D, 6C

transport.codes <- c('MEDT: MEDIC TRANSPORT', 'MEDIC TR RESP & TRANSPORTED')
cin1$med_transport <-cin1$disposition_text %in% transport.codes

cin1$breath.transport<-cin1$breath2 * cin1$med_transport

cin1$flu.transport<-cin1$flu * cin1$med_transport

```

```{r}
cin2<-cin1[,c('date','breath2', 'flu.transport','breath.transport','incident_type_id')]
cin2$state <- 'OH'
cin2$age <- 1
```

Majority of flu calls are for 36A-3-S, flu symptoms, surveillance only only. 36D are most severe

```{r}
sort(-table(cin1$incident_type_id[cin1$flu==1]))
sort(-table(cin1$incident_type_id[cin1$flu.transport==1]))

```
Pandemic/epidemic codes
36-D-1 Ineffective breathing with flu symptoms
36-D-2 Difficulty speaking between breaths with flu symptoms
36-D-3 Not alert with flu symptoms
36-D-4 Changing color with flu symptoms
36-C-1 Abnormal breathing with single flu symptom
36-C-2 Abnormal breathing with multiple flu symptom
36-C-3 Chest pain/discomfort ≥35 with single flu symptom
36-C-4 Chest pain/discomfort ≥35 with multiple flu symptoms
36-C-5 High risk conditions
36-A-1 Chest pain/discomfort <35 with single flu symptom
36-A-2 Chest pain/discomfort <35 with multiple flu symptoms
36-A-3 Flu symptoms only
```{r}
cin2$breath.flu.transport <- 0
cin2$breath.flu.transport[substr(cin2$incident_type_id,1,4) %in% c('36D1','36D2','36C1','36C2')] <- 1

cin2$all.breath.transport<-cin2$breath.flu.transport +cin2$breath.transport
```


Note: there is a flag for flu, but it only started being used in March 2020
```{r}
ts1<-ts_format(line.list=cin2, datevar='date', statevar='state',agevar='age', syndromes=c('breath2','breath.transport','breath.flu.transport','all.breath.transport'), remove.final = T)
```

Breathing problems, transported to hospital
```{r}
plot(ts1$date[1811:1911], ts1$breath.transport[1811:1911], type='l')
```

New flu code
```{r}
plot(ts1$date[1811:1911], ts1$breath.flu.transport[1811:1911], type='l')
```

#Breathing + new flu/breathing codes
```{r}
plot(ts1$date[1811:1911], ts1$all.breath.transport[1811:1911], type='l')
```
```{r}
excess_cases1 <-
  excessCases(ds = ts1,
              datevar = "date",
              agevar = "age",
              statevar = "state",
              denom.var = "all.visits",
              use.syndromes = c("all.breath.transport"),
              time.res = "day",
              flu.import=T, 
              rsv.import=F)
```



```{r}
dashboardPlot(excess_cases1)
```

## Santa monica
```{r}

url.sm <- "https://data.smgov.net/resource/5y3u-5db4.json"
sm1 <- read.socrata(url.sm) 
sm2 <- sm1[sm1$call_type_description =='Emergency Medical Service (EMS)',]

sm2$state <- "CA"
sm2$agevar <- '1'
sm2$ems <- 1

sm3 <- ts_format(line.list=sm2, datevar="incident_date", statevar='state',agevar='agevar', syndromes=c('ems'), remove.final = T)
sm3$all.visits <- 1

excess_cases.sm <-
  excessCases(ds = sm3,
              datevar = "incident_date",
              agevar = "agevar",
              statevar = "state",
              denom.var = "all.visits",
              use.syndromes = c("ems"),
              time.res = "day",
              flu.import=T, 
              rsv.import=F)
```

```{r}
dashboardPlot(excess_cases.sm)

```




