---
title: "nyc_syndromic_daily"
author: "Dan Weinberger, kelsie Cassell"
date: "3/10/2020 - KC edits:3/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(lubridate)
# library(INLA)
library(MMWRweek)
library(cdcfluview)
library(gtrendsR)
library(HDInterval)
library(shiny)
library(pbapply)
library(zoo)
library(ciTools)
library(MASS)
library(tidyr)
library(rsatscan)
library(readxl)
```

last date for which data are available (as specificed on folder name of data) YYYY_MM_DD
```{r, setdate}
last.date<-'2020_03_18'
```


##Import and clean the data
```{r}
ili1<- read.table(paste0('../Data/nyc_update_',last.date,'/ili_count.csv'),sep="\t", header=TRUE, skipNul = T, fileEncoding="UCS-2LE") # had to add "fileEncoding="UCS-2LE"" 
ili1$date<-as.Date(ili1$Date, '%m/%d/%y')
names(ili1)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'ili', 'ddate')
ili1<-ili1[,c('ddate', 'borough','age_group', 'ili')]

ili1.prop<- read.table(paste0('../Data/nyc_update_',last.date,'/ili_prop.csv'),sep="\t", header=TRUE, skipNul = T, fileEncoding="UCS-2LE") # had to add "fileEncoding="UCS-2LE"" 
ili1.prop$date<-as.Date(ili1.prop$Date, '%m/%d/%y')
names(ili1.prop)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'ili.prop', 'ddate')
ili1.prop<-ili1.prop[,c('ddate', 'borough','age_group', 'ili.prop')]

resp1<-read.table(paste0('../Data/nyc_update_',last.date,'/resp_count.csv'),sep="\t", header=TRUE, skipNul = T, fileEncoding="UCS-2LE") 
resp1$date<-as.Date(resp1$Date, '%m/%d/%y')
names(resp1)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'resp', 'ddate')
resp1<-resp1[,c('ddate', 'borough','age_group', 'resp')]

resp1.prop<-read.table(paste0('../Data/nyc_update_',last.date,'/resp_prop.csv'),sep="\t", header=TRUE, skipNul = T, fileEncoding="UCS-2LE") 
resp1.prop$date<-as.Date(resp1.prop$Date, '%m/%d/%y')
names(resp1.prop)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'resp.prop', 'ddate')
resp1.prop<-resp1.prop[,c('ddate', 'borough','age_group', 'resp.prop')]

ds.list<-list(resp1,resp1.prop, ili1, ili1.prop)
combo1 <- Reduce(
  function(x, y, ...) merge(x, y,by = c('ddate', 'borough', 'age_group'), all = TRUE, ...),
  ds.list
)
combo1 = combo1[,c('ddate', 'borough','age_group', 'ili', 'resp', 'ili.prop', 'resp.prop')]
combo1$ili<-gsub(',', '', combo1$ili) #remove commas from number
combo1$resp<-gsub(',', '', combo1$resp) #remove commas from number
combo1$ili = as.numeric(as.character(combo1$ili))
combo1$resp = as.numeric(as.character(combo1$resp))
combo1$ili.prop = as.numeric(as.character(combo1$ili.prop))
combo1$resp.prop = as.numeric(as.character(combo1$resp.prop))
combo1$resp.denom<-combo1$resp/combo1$resp.prop
combo1$ili.denom<-combo1$ili/combo1$ili.prop
combo1$denom<-apply(cbind(combo1$resp.denom,combo1$ili.denom),1,mean, na.rm=T)

combo1$agec[combo1$age_group == 'Ages 0-4 years']<-1
combo1$agec[combo1$age_group == 'Ages 5-17 years']<-2
combo1$agec[combo1$age_group == 'Ages 18-64 years']<-3
combo1$agec[combo1$age_group == 'Ages 65+ years']<-4
combo1$agec[combo1$age_group == 'All age groups']<-5 # here is all age groups! 
combo1<-combo1[combo1$agec %in% c(1,2,3,4,5),]
#Esnure all dates represented in the data
combo1<-complete(combo1, ddate = seq.Date(from=min(combo1$ddate), to=max(combo1$ddate), by='day'))
combo1<-combo1[,c('ddate', 'borough', 'agec','ili','resp','denom', 'ili.prop','resp.prop')]
saveRDS(combo1, '../Data/NYC_resp_ili_combo.rds')

```


```{r}
#combo1<-readRDS( '../Data/NYC_resp_ili_combo.rds')

```

```{r}

ili.df = combo1[,c('ddate','borough', 'agec', 'ili')]
names(ili.df) = c('ddate','borough', 'agec', 'value')
ili.df$variable = 'ili'
ili.df$value = ili.df$value

resp.df = combo1[,c('ddate','borough', 'agec', 'resp')]
names(resp.df) = c('ddate','borough', 'agec', 'value')
resp.df$variable = 'resp'
resp.df$value = resp.df$value

combo1.m<-melt(combo1[,c('ddate','borough', 'agec','ili','resp','denom')], id.vars = c('ddate','borough', 'agec'))
ds1.c<-acast(combo1.m,agec~ddate ~borough~variable)
#ds1.c[is.na(ds1.c)]<-0
#saveRDS(ds1.c,'../Data/nyc_shiny_data.rds') #save data for the shiny app

```

## View data

```{r, eval=F}
boroughs<-dimnames(ds1.c)[[3]]
boroughs<-boroughs[boroughs!='NA']
ages<-dimnames(ds1.c)[[1]]
ages<-ages[ages %in% c('1','2','3','4')]
n.times<-dim(ds1.c)[2]
Age.labels<-c('<5y','5-17y','18-64y','65+y', 'All ages')
#syndromes<- dimnames(ds1.c)[[4]]
syndromes<-c('ili','resp') # can also plot 'resp'
for(j in syndromes){
  for(i in boroughs){
    par(mfrow=c(2,2))  
    for(k in ages){
      plot(ds1.c[k,((n.times-365):n.times),i,j], type='l', bty='l', main=paste(j,i,Age.labels[as.numeric(k)] ) )
    }
  }
}
```




##Import auxillary data
Pulls in google searches for RSV in CT can get 5 years of historical data at weekly reslution (switches to monthly for >5 years)
```{r}
rsv.down<-gtrends(keyword = 'RSV', geo="US-NY-501", time = "today+5-y",
                  gprop = c("web") , category = 0,  low_search_volume = FALSE)
saveRDS(rsv.down,'../rsv_searches_nyc.rds')
# saveRDS(rsv.down,'../Data/rsv_searches_nyc.rds')
rsv<-rsv.down$interest_over_time[,c('date','hits')]
rsv$date<-as.Date(rsv$date)
mmwr.week.rsv<-MMWRweek(rsv$date)[,c('MMWRyear','MMWRweek')]
rsv<-cbind.data.frame(rsv,mmwr.week.rsv)
names(rsv)<-c('date','rsv.searches','MMWRyear','MMWRweek')
```

flusurvnet data
https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html

```{r}
flu<-read.csv('../Data/FluSurveillance_EIP_New York - Albany_Data.csv', skip=2)
flu2<-flu[flu$AGE.CATEGORY %in% c('Overall'),]
flu2<-flu2[,c('SEASON', 'MMWR.YEAR', 'MMWR.WEEK', 'WEEKLY.RATE')]
names(flu2)<-c('season', 'MMWRyear', 'MMWRweek','flu.rate')
flu2$flu.rate<-as.numeric(as.character(flu2$flu.rate))
flu2$flu.rate[is.na(flu2$flu.rate)]<-9999
```

Aggregate by week and merge auxillary variables
```{r}
combo1 = readRDS('../Data/NYC_resp_ili_combo.rds')

mmwr.date<-MMWRweek(combo1$ddate)
ds1.df<-cbind.data.frame(combo1,mmwr.date)
combo2<-merge(ds1.df, flu2, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2<-merge(combo2, rsv, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2$flu.rate[is.na(combo2$flu.rate)]<-0

#ensures flu variable set to missing if no flu observations
combo2$flu.rate[combo2$flu.rate==9999]<-NA #last few weeks
max.date.miss.flu<-max(combo2$ddate[is.na(combo2$flu.rate)])
combo2$flu.rate[combo2$ddate>max.date.miss.flu] <-NA

combo2.sub<-combo2[, c('agec', 'ddate','MMWRyear', 'MMWRweek', 'borough', 'ili','resp','flu.rate',"rsv.searches", 'ili.prop','resp.prop')]
ili.m<-melt(combo2.sub, id.vars=c('borough','agec','ddate','MMWRyear','MMWRweek'))
ili.a<-acast(ili.m, ddate+MMWRyear+MMWRweek ~borough~agec~variable , fun.aggregate = sum )
dimnames(ili.a)[[1]]<-substr(dimnames(ili.a)[[1]],1,10)

# ili.a.df = as.data.frame(ili.a)
```


setup epiyear index
```{r}
date.string<-as.Date(dimnames(ili.a)[[1]])
month<-month(date.string)
epiyr<-year(date.string)
epiyr[month<=6] <- epiyr[month<=6]-1
epiyr.index<-epiyr-min(epiyr)+1
weekN<-MMWRweek(date.string)[,'MMWRweek']
day.of.year<-yday(date.string)
day.of.week<-as.factor(weekdays(date.string))
```

## Evaluate results after controlling for flu and RSV
```{r}
glm.func<-function(x.test, age.test, syndrome){
  clean.array.citywide<-ili.a[,x.test,,]
  epiyr.index.f<-as.factor(epiyr.index)
  epiyr.index.f2<-as.factor(epiyr.index)
  
    y.age = t(clean.array.citywide[,age.test,syndrome])
      n.dates<-length(y.age)
    
      if(syndrome=='ili'){
        denom<-y.age[1,]/t(clean.array.citywide[,age.test,'ili.prop'])[1,]}else{
       denom<-y.age[1,]/t(clean.array.citywide[,age.test,'resp.prop']+0.01)[1,]
                  }

      
  y.age.fit<-y.age[1,]
  y.age.fit[date.string>=as.Date("2020-03-01")] <- NA #extrapolate last 1 months
  
  sqrt.rsv =sqrt(clean.array.citywide[,age.test,'rsv.searches']) #same for all ages and boroughs
  sqrt.flu =sqrt(clean.array.citywide[,age.test,'flu.rate']) #same for all ages and boroughs
  
  
  sqrt.flu<- na.locf(sqrt.flu)  #fill in missing observations for flu at end of TS with most recent observed values
  
    sqrt.rsv<- na.locf(sqrt.rsv)  #fill in missing observations for RSV at end of TS with most recent observed values

    
  t2<-1:length(y.age)
  # 
  sin1<-sin(2*pi*t2/365.25)
  cos1<-cos(2*pi*t2/365.25)
  sin2<-sin(2*pi*t2*2/365.25)
  cos2<-cos(2*pi*t2*2/365.25)
  sin3<-sin(2*pi*t2*3/365.25)
  cos3<-cos(2*pi*t2*3/365.25)
  
  
  ds.glm<-cbind.data.frame('day.of.year'=day.of.year,'y.age'=y.age[1,],'y.age.fit'=y.age.fit,sqrt.rsv, sqrt.flu, day.of.week, t2,epiyr.index.f, sin1, sin2, cos1, cos2, sin3, cos3)
  #ds.glm<-ds.glm[complete.cases(ds.glm),]

  form1<-as.formula(paste0('y.age.fit',"~ epiyr.index.f*sqrt.rsv +   #rsv effect varies by epiyr
                   epiyr.index.f*sqrt.flu + #flu effect, varies by epiyear
                   day.of.week+
                   sin1+cos1 +sin2+cos2+ sin3+cos3 "
  ))
  form2<-as.formula(paste0('y.age',"~ epiyr.index.f*sqrt.rsv +   #rsv effect varies by epiyr
                   epiyr.index.f*sqrt.flu + #flu effect, varies by epiyear
                   day.of.week+
                   sin1+cos1 +sin2+cos2+ sin3+cos3  "
  ))
  mod1<-glm(form1, data=ds.glm, family=poisson(link='log'))
  #pred1<-predict(mod1, type='response', newdata = ds.glm)
  #my_data_with_pi <- add_pi(ds.glm, mod1, names = c("lpb", "upb"))

  #500 samples total
   pred.coefs.reg.mean<- mvrnorm(n = 100, mu=coef(mod1), Sigma=vcov( mod1))
   mod.mat.pred<-model.matrix(form2, data=ds.glm, family='poisson')
preds.stage1.regmean<- mod.mat.pred %*% t(pred.coefs.reg.mean) 
preds.stage2<-rpois(n=length(preds.stage1.regmean)*5, exp(preds.stage1.regmean))
preds.stage2<-matrix(preds.stage2, nrow=nrow(preds.stage1.regmean), ncol=ncol(preds.stage1.regmean)*5)
 preds.stage2.q<-t(apply(preds.stage2,1,quantile, probs=c(0.025,0.5, 0.975)))                   
  
   resid1<- log((ds.glm$y.age+0.5) /(preds.stage2.q[,'50%']+0.5))

  
  out.list<-list(y=y.age[1,], pred=preds.stage2.q[,'50%'], resid1=resid1, upi=preds.stage2.q[,'97.5%'], lpi=preds.stage2.q[,'2.5%'],'ili.prop'= t(clean.array.citywide[,age.test,'ili.prop']), 'resp.prop'= t(clean.array.citywide[,age.test,'resp.prop'])
 )
  return(out.list)
}

syndromes<- c('ili', 'resp') 
ages <-   c('1','2','3','4','5')
counties.to.test<-c('Bronx','Brooklyn','Manhattan','Queens','Staten Island','Citywide')

all.glm.res<- pblapply(syndromes, function(x){
  ww<- lapply(ages, function(y){
    lapply(counties.to.test, glm.func, age.test=y, syndrome=x)
    }
  ) 
  names(ww)<- ages
  return(ww)
}
)
names(all.glm.res)<-syndromes
saveRDS(all.glm.res, './nyc_shiny_data/glm.results.rds')
```

##When updating app, can stop here
#######################################
######################################


Next extract residuals and fitted values
```{r, fig.width=8, fig.height=6}
#Residuals for all age groups for ILI
ili2.resid<- sapply(all.glm.res[['ili']], function(x) sapply(x,'[[','resid1'), simplify='array')
dimnames(ili2.resid)[[2]]<-counties.to.test
ili2.pred<- sapply(all.glm.res[['ili']], function(x) sapply(x,'[[','pred'), simplify='array')
dimnames(ili2.pred)[[2]]<-counties.to.test
ili2.pred.lcl<- sapply(all.glm.res[['ili']], function(x) sapply(x,'[[','lpi'), simplify='array')
dimnames(ili2.pred.lcl)[[2]]<-counties.to.test
ili2.pred.ucl<- sapply(all.glm.res[['ili']], function(x) sapply(x,'[[','upi'), simplify='array')
dimnames(ili2.pred.ucl)[[2]]<-counties.to.test
obs.ili<- sapply(all.glm.res[['ili']], function(x) sapply(x,'[[','y'), simplify='array')

dimnames(obs.ili)[[2]]<-counties.to.test
dates<-as.Date(dimnames(ili.a)[[1]])
```


Observed vs expected by county and age group
```{r}
age.labels = c("Ages 0-4 years", "Ages 5-17 years", "Ages 18-64 years", "Ages 65+ years", "All age groups")

n.times<-length(dates)
dates.select<-(n.times-360):n.times
par(mfrow=c(2,3), mar=c(3,2,1,1))
for(i in 1:5){
  for( j in counties.to.test){
  plot(dates[dates.select],obs.ili[dates.select,j,i], type='l', bty='l', ylab='Fitted', main=paste(j, age.labels[i]), ylim=range(c(ili2.pred.lcl[dates.select,j,i],ili2.pred.ucl[dates.select,j,i],ili2.pred[dates.select,j,i]), na.rm=T ))
  points(dates[dates.select],ili2.pred[dates.select,j,i], type='l', col='red', lty=3 )
  points(dates[dates.select],ili2.pred.lcl[dates.select,j,i], type='l', col='red', lty=3 )
  points(dates[dates.select],ili2.pred.ucl[dates.select,j,i], type='l', col='red', lty=3 )
  abline(v=790)
  }
}
pdf(file='../Results/nyc_borough_age.pdf', width=10, height=5)
par(mfrow=c(2,3), mar=c(3,2,2,1))
for(i in 1:5){
  for( j in counties.to.test){
  plot(dates[dates.select],obs.ili[dates.select,j,i], type='l', bty='l', ylab='Fitted', main=paste(j, age.labels[i]), ylim=range(c(ili2.pred.lcl[dates.select,j,i],ili2.pred.ucl[dates.select,j,i],ili2.pred[dates.select,j,i]), na.rm=T ))
  points(dates[dates.select],ili2.pred[dates.select,j,i], type='l', col='red', lty=2, lwd=2 )
  points(dates[dates.select],ili2.pred.lcl[dates.select,j,i], type='l', col='red', lty=3 , lwd=2)
  points(dates[dates.select],ili2.pred.ucl[dates.select,j,i], type='l', col='red', lty=3 ,lwd=2)
  abline(v=790)
  }
}
dev.off()
```




Pop size file by borough
```{r}
pop1<-read.csv('../Data/borough_pop.csv')
pop1$agec<-NA
pop1$agec[pop1$age %in% c('0-4')]<-1
pop1$agec[pop1$age %in% c('5-9','10-14','15-19')]<-2
pop1$agec[pop1$age %in% c('20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64')]<-3
pop1$agec[pop1$age %in% c('65-69', '70-74','75-79','80-84','85+')]<-4

pop1.cw<-pop1
pop1.cw$borough<-'Citywide'
pop1.cw<-aggregate(pop1.cw[,c('X2015','X2020')],by=list('agec'=pop1.cw$agec, 'borough'=pop1.cw$borough), FUN=sum)

pop2<-aggregate(pop1[,c('X2015','X2020')], by=list('agec'=pop1$agec, 'borough'=pop1$borough), FUN=sum)
pop1$agec<-5
pop2.all<-aggregate(pop1[,c('X2015','X2020')],by=list('agec'=pop1$agec,'borough'=pop1$borough), FUN=sum)
pop2.all.cw<-pop1
pop2.all.cw$borough<-'Citywide'
pop2.all.cw<-aggregate(pop2.all.cw[,c('X2015','X2020')],by=list('agec'=pop2.all.cw$agec,'borough'=pop2.all.cw$borough), FUN=sum)




pop3.all<-rbind.data.frame(pop2,pop2.all,pop2.all.cw,pop1.cw)
names(pop3.all)<-c('agec','borough', '2015-01-01','2020-01-01')

all.dates<-as.character(date.string)
all.dates[!all.dates %in% c('2015-01-01', '2020-01-01')]
all.ages<-unique(pop3.all$agec)
all.boroughs<-unique(pop3.all$borough)
all.combos<-expand.grid(list('agec'=all.ages,'variable'=all.dates, 'borough'=all.boroughs))
all.combos$value<-NA

pop3.all.m<-melt(pop3.all, id.vars=c('agec', 'borough'))
pop3.all.m$variable<-as.character(as.numeric(pop3.all.m$variable))
pop3.all.m<-rbind.data.frame(pop3.all.m,all.combos)
pop3.all.c<-acast(pop3.all.m, variable~borough~agec)
#names(pop3.all.c)<-c('agec','year','pop')
pop3<-apply(pop3.all.c, c(2,3), function(z){
  interpol1<-approx(z, n=length(date.string))
  return(interpol1[[2]])
}
  )
names(pop3)<-c('date','borough','agec')
dimnames(pop3)[[1]]<-date.string
simpleCap <- function(x) {
  s <- strsplit(x, " ")[[1]]
  paste(toupper(substring(s, 1,1)), substring(s, 2),
      sep="", collapse=" ")
}
dimnames(pop3)[[2]]<-sapply(dimnames(pop3)[[2]], simpleCap)
saveRDS(pop3,'./nyc_shiny_data/borough.pop.rds')
```


## Temporal SATSCAN by county and syndrome
Do a Poisson temporal satscan, adjusting for fitted values as the population file
-NOTE: this ONLY works if you have SATSCAN installed on your machine
```{r}
syndromes<- c('ili', 'resp') 
ages <-   c('1','2','3','4','5')
counties.to.test<-dimnames(ili.a)[[2]]
dates<-as.Date(dimnames(ili.a)[[1]])
satscan.fun<-function(syndrome, county.select,age.select){
  obs<- sapply(all.glm.res[[syndrome]], function(x) sapply(x,'[[','y'), simplify='array')
  dimnames(obs)[[2]]<-counties.to.test
  
  case.pred<- sapply(all.glm.res[[syndrome]], function(x) sapply(x,'[[','pred'), simplify='array')
  dimnames(case.pred)[[2]]<-counties.to.test
  
  mycas = data.frame(locid=1:dim(case.pred)[[1]] ,basecas=obs[,county.select,age.select], day=dates)
  mypop<- data.frame(locid=1:dim(case.pred)[[1]] ,day=dates,basepop=case.pred[,county.select,age.select] )
  mygeo<- data.frame(locid=1:dim(case.pred)[[1]] ,lat=1:dim(case.pred)[[1]], long=1:dim(case.pred)[[1]])
  
  #Write satscan files
  
  td = tempdir()
  write.cas(mycas, location = td, file = "mycas")
  write.pop(mypop, location = td, file = "mypop")
  write.geo(mygeo, location = td, file = "mygeo")
  
  #write.geo(mygeo, location = '../satscan/data/', file = "geo.spacetime.ct", userownames=F)
  #write.cas(mycas, location = '../satscan/data/', file = "ili.temporal")
  #write.pop(mypop, location = '../satscan/data/', file = "ili.expected")
  
  #Parameter files
  
  first.date<-as.character(min(mycas$day))
  first.date<-gsub('-', '/',first.date)
  last.date<-as.character(max(mycas$day))
  last.date<-gsub('-', '/',last.date)
  invisible(ss.options(reset=TRUE))
  ss.options(list(CaseFile="mycas.cas",PopulationFile='mypop.pop', CoordinatesFile="mygeo.geo", PrecisionCaseTimes=3))
  ss.options(list(StartDate=first.date, CoordinatesType=1, TimeAggregationUnits=3,TimeAggregationLength=5))
  ss.options(list(EndDate=last.date,  AnalysisType=6, ModelType=0,ScanAreas=1)) 
  ss.options(list(ProspectiveStartDate="2019/01/01", ReportGiniClusters="n", LogRunToHistoryFile="n"))
  
  write.ss.prm(td, "mybase")
  #write.ss.prm('../satscan/satscan parameters', "temporal adjust rsatscan")
  
  mybase = satscan(td, "mybase")
  return(mybase)
}


all.satscan.res<- pblapply(syndromes, function(x){
  ww<- lapply(ages, function(y){
    lapply(counties.to.test, satscan.fun, age.select=y, syndrome=x)
  }
  ) 
  names(ww)<- ages
  return(ww)
}
)
names(all.satscan.res)<-syndromes
```
# Results from SATSCAN

```{r}
#Extract results
ili.satscan<- sapply(all.satscan.res[['ili']], function(x) sapply(x,'[[','col', simplify=F), simplify='F')

all.ages.ili<-lapply(all.satscan.res[['ili']], function(y) sapply(y, function(x) apply(x$col,2,function(z) as.character(z))))

all.ages.ili<-lapply(all.ages.ili, function(x){ 
  colnames(x)=counties.to.test
  return(x)})
names(all.ages.ili)<-age.labels[1:length(all.ages.ili)]
all.ages.ili

```


```{r, fig.width=12, fig.height=8}
diff.resid<-ili2.resid[,,'4'] - ili2.resid[,,'1']
par(mfrow=c(2,4))
lapply(counties.to.test, function(x){
  plot(dates[dates>=as.Date("2020-01-01")],diff.resid[dates>=as.Date("2020-01-01"),x], type='l', bty='l', main=x)
  abline(h=0, lty=2, col='gray')
  })
```
```