---
title: "covid_19_syndromic"
author: "Dan Weinberger"
date: "3/8/2020"
output: html_document
---

The goal of these analyses is to detect increases in syndromic disease activity associated with  COVID-19. As a test, we are using syndromic disease data from New York City. Data are stratified by day, borrough, age group. the data were downloaded from the [Epiquery website](https://a816-health.nyc.gov/hdi/epiquery/visualizations?PageType=ps&PopulationSource=Syndromic)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("..")) #sets Working directory to be 1 folder up

library(reshape2)
library(lubridate)
library(mgcv)
library(INLA)
library(inlabru)
library(MMWRweek)
library(cdcfluview)
library(gtrendsR)
library(rjags)
library(HDInterval)
```

## Read in the data

Import the data and remove the unnecesary rows. Note that the file downloads as a .csv format, but it is actually a tab-separated file.

This pulls in the ILI syndromic data from Epiquery in NYC
```{r, eval=F}
ili1<- read.table('./Data/ILI_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 
resp1<-read.table('./Data/resp_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 
combo1<-rbind.data.frame(ili1, resp1)

names(combo1)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'count')

combo1<-combo1[,c('syndrome','borough','age_group','date','count')]
combo1$date<-as.Date(combo1$date, '%m/%d/%y')
mmwr.week<-MMWRweek(combo1$date)
combo1<-cbind.data.frame(combo1, mmwr.week)
combo1$count<-gsub(',', '', combo1$count) #remove commas from number
combo1$count<-as.numeric(as.character(combo1$count))

saveRDS(combo1,'./Data/ili_resp.rds')
```


```{r}
combo1<-readRDS('./Data/ili_resp.rds')
```

This pulls in the NREVSS syndromic data from CDC (not currently being used)
```{r}
flu<-read.csv('./Data/FluSurveillance_EIP_New York - Albany_Data.csv', skip=2)
#nrevss<-who_nrevss(region=c('state'))
#clin.ny<-nrevss$clinical_labs
#clin.ny<-clin.ny[clin.ny$region=='New York',]
#clin.ny$percent_positive<-as.numeric(clin.ny$percent_positive)
#clin.ny$wk_date<-NULL
#clin.ny[,-c(1:2)]<-apply(clin.ny[,-c(1:2)],2, function(x) as.numeric(x))
#flu<-clin.ny
#ph.ny<-nrevss$public_health_labs
#ph.ny<-ph.ny[ph.ny$region=='New York',]
```

Pulls in google searches for RSV in NYC. can get 5 years of historical data at weekly reslution (switches to monthly for >5 years)
```{r}
rsv.down<-gtrends(keyword = 'RSV', geo="US-NY-501", time = "today+5-y",
gprop = c("web") , category = 0,  low_search_volume = FALSE)
saveRDS(rsv.down,'./Data/rsv_searches_nyc.rds')
rsv<-rsv.down$interest_over_time[,c('date','hits')]
rsv$date<-as.Date(rsv$date)
mmwr.week.rsv<-MMWRweek(rsv$date)[,c('MMWRyear','MMWRweek')]
rsv<-cbind.data.frame(rsv,mmwr.week.rsv)
names(rsv)<-c('date','rsv.searches','MMWRyear','MMWRweek')
```


```{r}
#select.syndrome<-"Respiratory"
select.syndrome<-"Influenza-like illness (ILI)"
```

## Reshape the ILI data 
Aggregate by MMWR date. For now, just focus on ILI
```{r}
combo1$MMWRday<-NULL
combo1$date<-NULL

combo1.m<-melt(combo1[combo1$syndrome==select.syndrome,], id.vars=c('syndrome', 'borough','age_group','MMWRyear','MMWRweek'))
combo1.a<-dcast(combo1.m, MMWRyear+MMWRweek +borough+syndrome~age_group, fun.aggregate = sum )
combo1.a$"o65"<-combo1.a$`Ages 65+ years`
combo1.a$"u5"<-combo1.a$`Ages 0-4 years`
combo1.a$epiyr<-combo1.a$MMWRyear
combo1.a$epiyr[combo1.a$MMWRweek<=26] <-combo1.a$epiyr[combo1.a$MMWRweek<=26] -1
combo1.a$date.index<- combo1.a$MMWRyear +combo1.a$MMWRweek/52-1/52
```

## Explore the city-wide data from NYC
```{r}
nyc<-combo1.a[combo1.a$borough=='Citywide',]
#nyc<-combo1.a[combo1.a$borough=='Manhattan',]

```

```{r}
plot(nyc$date.index,nyc$"Ages 18-64 years", type='l', bty='l', main='ILI all ages in NYC')
```

```{r, fig.width=4, fig.height=6}
par(mfrow=c(2,1), mar=c(2,2,2,1))
plot(nyc$date.index,nyc$`Ages 0-4 years`, type='l', bty='l', main='ILI <5 years in NYC')
plot(nyc$date.index,nyc$`Ages 65+ years`, type='l', bty='l', main='ILI 65+ years in NYC')
```

Plot ratio of 65+ to younger age groups for ILI
```{r, fig.width=12, fig.height=4}
ages<-c("Ages 0-4 years", "Ages 5-17 years",  "Ages 18-64 years" )
plot.rat<-function(age.select){
nyc$age.rat<-(nyc$`Ages 65+ years`+0.5)/(nyc[,age.select]+0.5)
plot(nyc$date.index,log(nyc$age.rat), type='l', bty='l', main=paste0('ratio of 65+ and ',age.select), yaxt='n', ylab='Ratio')
axis(side=2, at=-c(-5, -0.5, 0, 0.5), label=round(exp(c(-5, -0.5, 0, 0.5)),2) )
}
par(mfrow=c(1,3))
lapply(ages, plot.rat)
```



Disaggregated boroughs data. 
```{r}
bur1<- combo1.a[combo1.a$borough != 'Citywide',]
bur1$borough<-factor(bur1$borough)
```



## Prepare data for JAGS
Create a 4D array with all of the data

```{r}
ili<-combo1[combo1$syndrome==select.syndrome & combo1$age_group %in% c('Ages 0-4 years','Ages 65+ years'),  c('age_group','syndrome', 'borough','MMWRyear','MMWRweek','count')]
names(ili)<-c('age_group','syndrome', 'borough','MMWRyear','MMWRweek','ili')
ili$agec<-NA
ili$agec[ili$age_group=='Ages 0-4 years']<-1
ili$agec[ili$age_group=='Ages 65+ years']<-5
```

Use overall aggregate flu hospitalizations
```{r}
flu2<-flu[flu$AGE.CATEGORY %in% c('Overall'),]
#flu2$agec<-NA
#flu2$agec[flu2$AGE.CATEGORY=='0-4 yr']<-1
#flu2$agec[flu2$AGE.CATEGORY=='65+ yr']<-5
flu2<-flu2[,c('SEASON', 'MMWR.YEAR', 'MMWR.WEEK', 'WEEKLY.RATE')]
names(flu2)<-c('season', 'MMWRyear', 'MMWRweek','flu.rate')
flu2$flu.rate<-as.numeric(as.character(flu2$flu.rate))
flu2$flu.rate[is.na(flu2$flu.rate)]<-9999

```

```{r}
combo2<-merge(ili, flu2, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2<-merge(combo2, rsv, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2$flu.rate[is.na(combo2$flu.rate)]<-0
combo2$flu.rate[combo2$flu.rate==9999]<-NA #last few weeks
```


```{r}
combo2.sub<-combo2[, c('agec', 'MMWRyear', 'MMWRweek', 'borough', 'ili','flu.rate',"rsv.searches")]
ili.m<-melt(combo2.sub, id.vars=c('borough','agec','MMWRyear','MMWRweek'))
ili.a<-acast(ili.m, MMWRyear+MMWRweek ~borough~agec~variable , fun.aggregate = sum )
clean.array.jags<-ili.a[,c('Bronx','Brooklyn','Manhattan','Queens' ,'Staten Island'),c('1','5'),]
clean.array.jags[216,,,"flu.rate"]<-clean.array.jags[215,,,"flu.rate"] #patch for now to deal with missing last week of 
saveRDS(clean.array.jags,'./Data/jags.ds.rds')
```

```{r}
plot(scale(ili.a[,'Citywide','1','ili']), col='blue')
points(scale(ili.a[,'Citywide','1','flu.rate']), col='red')
points(scale(ili.a[,'Citywide','1','rsv.searches']), col='gray')

#Cross correlation
ccf(ili.a[,'Citywide','1','ili'], ili.a[,'Citywide','1',"rsv.searches"], na.action=na.omit)
```


## simple gam
```{r}
clean.array.citywide<-ili.a[,c('Citywide'),c('1','5'),]

y.kids = t(clean.array.citywide[,'1','ili'])
y.adult = t(clean.array.citywide[,'5','ili'])
        sqrt.rsv =sqrt(clean.array.citywide[,'5','rsv.searches']) #same for all ages and boroughs
                 sqrt.flu =sqrt(clean.array.citywide[,'5','flu.rate']) #same for all ages and boroughs
                 sqrt.flu[216]<-sqrt.flu[215]
t<-1:length(y.kids)
weekN<-rep(1:52, length.out=length(y.kids))

ds.glm<-cbind.data.frame('weekN'=weekN,'y.kids'=y.kids[1,],'y.adult'=y.adult[1,],sqrt.rsv, sqrt.flu, t)
mod1<-gam(y.adult~ sqrt.rsv+sqrt.flu +s(t,bs='bs') +s(weekN, bs='cc') , family='poisson', data=ds.glm)
  plot(mod1) 
pp1 <- predict(mod1, type = "terms")

mod2<-gam(y.kids~ sqrt.rsv+sqrt.flu +s(t, bs='bs') +s(weekN, bs='cc') , family='poisson', data=ds.glm)
  plot(mod2) 
pp2 <- predict(mod2, type = "terms")

diff1_2<- (pp2[,'s(t)']+ pp2[,'s(weekN)']) -(pp1[,'s(t)'] + + pp2[,'s(weekN)'])
  plot(diff1_2)
```


```{r}
date.string<-dimnames(ili.a)[[1]]
dates<-unlist(strsplit(date.string, '_'))
dates<-matrix(as.numeric(dates), ncol=2, byrow=T)
epiyr<-dates[,1]
epiyr[dates[,2]<=26] <- epiyr[dates[,2]<=26]-1
epiyr.index<-epiyr-min(epiyr)+1
```


## Stage 1 model: 
Estimate variation in ILI that is not associated with RSv or flu
```{r}
###############
#JAGS Model
###############
model_string<-"

model{

for(i in 1:n.space){

   for(t in 1:n.time){

      y.kid[i,t] ~ dpois(lambda.kid[i,t])
      log(lambda.kid[i,t]) <- alpha1[i] + epsilon[epiyr.index[i],1,i]*sqrt.flu[t] + epsilon[epiyr.index[i],2,i]*sqrt.rsv[t] + non.flu.kid[i,t]
    
      y.adult[i,t] ~ dpois(lambda.adult[i,t])
      log(lambda.adult[i,t]) <- alpha2[i] + delta[epiyr.index[i],1,i]*sqrt.flu[t] + delta[epiyr.index[i],2,i]*sqrt.rsv[t] + non.flu.adult[i,t]
    
    # We are interested in CHANGES in dufference between non.flu.adult[i,t] and non.flu.kid[i,t]
      
      diff_hat[i,t] <- non.flu.adult[i,t] - non.flu.kid[i,t] #difference bc on log scale
      }
}

for(i in 1:n.space){
  non.flu.kid[i,1]   <- 0 #t=0 is captured by intercept
  non.flu.adult[i,1] <- 0
for(t in 2:n.time){
  non.flu.kid[i,t]~dnorm(non.flu.kid[i,t-1], tau.kid) #note shared tau for all time series
  non.flu.adult[i,t]~dnorm(non.flu.adult[i,t-1], tau.adult) #note shared tau for all time series
}
}

for(j in 1:n.space){
alpha1[j] ~ dnorm(0,1e-4)
alpha2[j] ~ dnorm(0, 1e-4)

for(k in 1:2){
for(m in 1:n.epiyr){
 epsilon[m,k,j]~ dnorm(mu.epsilon[k], tau.epsilon[k])
 delta[m,k,j]~dnorm(mu.delta[k], tau.delta[k])
  } 
 }
}

for(k in 1:2){
  mu.epsilon[k]~dnorm(0, 1e-4)
  mu.delta[k]~dnorm(0, 1e-4)
  tau.delta[k]<-1/sd.delta[k]^2
  tau.epsilon[k]<-1/sd.epsilon[k]^2
  sd.delta[k]~dunif(0, 100)
  sd.epsilon[k]~dunif(0, 100)
}



tau.kid<-1/sd.kid^2
sd.kid~dunif(0,100)
tau.adult<-1/sd.adult^2
sd.adult~dunif(0,100)

beta0 ~ dnorm(0.00, 1e-4)
beta1 ~ dnorm(0.00, 1e-4)
}
"

```

Call model
```{r}
  ##############################################################
  #Model Fitting
  ##############################################################
  inits1=list(".RNG.seed"=c(123), ".RNG.name"='base::Wichmann-Hill')
  inits2=list(".RNG.seed"=c(456), ".RNG.name"='base::Wichmann-Hill')
  inits3=list(".RNG.seed"=c(789), ".RNG.name"='base::Wichmann-Hill')
  burnN=5000
  sampleN=10000
  ##############################################
  #Model Organization
  ############################################
  

  model_spec<-textConnection(model_string)
  model_jags<-jags.model(model_spec, 
                         inits=list(inits1,inits2, inits3),
                         data=list('y.kid' = t(clean.array.jags[,,'1','ili']),
                                   'y.adult' = t(clean.array.jags[,,'5','ili']),
                                   'sqrt.rsv' =sqrt(clean.array.jags[,'Bronx','5','rsv.searches']), #same for all ages and boroughs
                                   'sqrt.flu' =sqrt(clean.array.jags[,'Bronx','5','flu.rate']), #same for all ages and boroughs
                                   'n.space'=dim(clean.array.jags)[2],
                                   'n.time'=dim(clean.array.jags)[1],
                                   'epiyr.index'=epiyr.index,
                                   'n.epiyr'=max(epiyr.index)
                                   ),
                         n.adapt=burnN, 
                         n.chains=3)
  params<-c('diff_hat','non.flu.adult','non.flu.kid')
  
  ##############################################
  #Posterior Sampling
  ##############################################
  posterior_samples<-coda.samples(model_jags, 
                                  params, 
                                  n.iter=sampleN)
  posterior_samples.all<-do.call(rbind,posterior_samples)
  post_means<-apply(posterior_samples.all, 2, median)
  sample.labs<-names(post_means)
  posterior_samples.mean<-apply(posterior_samples.all,2,mean)
  posterior_samples.sd<-apply(posterior_samples.all,2,sd)

  diff_hat_mean<-posterior_samples.mean[grep('diff_hat', names(posterior_samples.mean))]
  diff_hat_mean<-matrix(diff_hat_mean,ncol=5, byrow=T)
  
  non.flu.kid.mean<-posterior_samples.mean[grep('non.flu.kid', names(posterior_samples.mean))]
    non.flu.kid.mean<-matrix(non.flu.kid.mean,ncol=5, byrow=T)

  non.flu.adult.mean<-posterior_samples.mean[grep('non.flu.adult', names(posterior_samples.mean))]
    non.flu.adult.mean<-matrix(non.flu.adult.mean,ncol=5, byrow=T)
    
```

##Stage 2 model:

