---
title: "covid_19_syndromic"
author: "Dan Weinberger"
date: "3/8/2020"
output: html_document
---

The goal of these analyses is to detect increases in syndromic disease activity associated with  COVID-19. As a test, we are using syndromic disease data from New York City. Data are stratified by day, borrough, age group. the data were downloaded from the [Epiquery website](https://a816-health.nyc.gov/hdi/epiquery/visualizations?PageType=ps&PopulationSource=Syndromic)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("..")) #sets Working directory to be 1 folder up

library(reshape2)
library(lubridate)
library(mgcv)
library(INLA)
library(inlabru)
library(MMWRweek)
library(cdcfluview)
library(gtrendsR)
```

Import the data and remove the unnecesary rows. Note that the file downloads as a .csv format, but it is actually a tab-separated file.
```{r, eval=F}
ili1<- read.table('./Data/ILI_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 
resp1<-read.table('./Data/resp_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 
combo1<-rbind.data.frame(ili1, resp1)

names(combo1)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'count')

combo1<-combo1[,c('syndrome','borough','age_group','date','count')]
combo1$date<-as.Date(combo1$date, '%m/%d/%y')
mmwr.week<-MMWRweek(combo1$date)
combo1<-cbind.data.frame(combo1, mmwr.week)
combo1$count<-gsub(',', '', combo1$count) #remove commas from number
combo1$count<-as.numeric(as.character(combo1$count))

saveRDS(combo1,'./Data/ili_resp.rds')
```


## Read in the data
```{r}
combo1<-readRDS('./Data/ili_resp.rds')
```

NREVSS flu data
```{r}
flu<-read.csv('./Data/FluSurveillance_EIP_New York - Albany_Data.csv', skip=2)
#nrevss<-who_nrevss(region=c('state'))
#clin.ny<-nrevss$clinical_labs
#clin.ny<-clin.ny[clin.ny$region=='New York',]
#clin.ny$percent_positive<-as.numeric(clin.ny$percent_positive)
#clin.ny$wk_date<-NULL
#clin.ny[,-c(1:2)]<-apply(clin.ny[,-c(1:2)],2, function(x) as.numeric(x))
#flu<-clin.ny
#ph.ny<-nrevss$public_health_labs
#ph.ny<-ph.ny[ph.ny$region=='New York',]
```

google searches for RSV in NYC. can get 5 years of historical data
```{r}
rsv.down<-gtrends(keyword = 'RSV', geo="US-NY-501", time = "today+5-y",
gprop = c("web") , category = 0,  low_search_volume = FALSE)
saveRDS(rsv.down,'./Data/rsv_searches_nyc.rds')
rsv<-rsv.down$interest_over_time[,c('date','hits')]
rsv$date<-as.Date(rsv$date)
mmwr.week.rsv<-MMWRweek(rsv$date)[,c('MMWRyear','MMWRweek')]
rsv<-cbind.data.frame(rsv,mmwr.week.rsv)
names(rsv)<-c('date','rsv.searches','MMWRyear','MMWRweek')
```


```{r}
#select.syndrome<-"Respiratory"
select.syndrome<-"Influenza-like illness (ILI)"
```

Reshape the ILI data, aggregate by MMWR date. For now, just focus on ILI
```{r}
combo1$MMWRday<-NULL
combo1$date<-NULL

combo1.m<-melt(combo1[combo1$syndrome==select.syndrome,], id.vars=c('syndrome', 'borough','age_group','MMWRyear','MMWRweek'))
combo1.a<-dcast(combo1.m, MMWRyear+MMWRweek +borough+syndrome~age_group, fun.aggregate = sum )
```



```{r}
combo1.a$"o65"<-combo1.a$`Ages 65+ years`
combo1.a$"u5"<-combo1.a$`Ages 0-4 years`
combo1.a$epiyr<-combo1.a$MMWRyear
combo1.a$epiyr[combo1.a$MMWRweek<=26] <-combo1.a$epiyr[combo1.a$MMWRweek<=26] -1
combo1.a$date.index<- combo1.a$MMWRyear +combo1.a$MMWRweek/52-1/52

```

```{r}
nyc<-combo1.a[combo1.a$borough=='Citywide',]
#nyc<-combo1.a[combo1.a$borough=='Manhattan',]
#nyc<-combo1.a[combo1.a$borough=='Queens',]
#nyc<-combo1.a[combo1.a$borough=='Bronx',]

```


```{r}
plot(nyc$date.index,nyc$"Ages 18-64 years", type='l', bty='l', main='ILI all ages in NYC')
```

```{r, fig.width=4, fig.height=6}
par(mfrow=c(2,1), mar=c(2,2,2,1))
plot(nyc$date.index,nyc$`Ages 0-4 years`, type='l', bty='l', main='ILI <5 years in NYC')
plot(nyc$date.index,nyc$`Ages 65+ years`, type='l', bty='l', main='ILI 65+ years in NYC')
```
Plot ratio of 65+ to younger age groups for ILI
```{r, fig.width=12, fig.height=4}
ages<-c("Ages 0-4 years", "Ages 5-17 years",  "Ages 18-64 years" )
plot.rat<-function(age.select){
nyc$age.rat<-(nyc$`Ages 65+ years`+0.5)/(nyc[,age.select]+0.5)
plot(nyc$date.index,log(nyc$age.rat), type='l', bty='l', main=paste0('ratio of 65+ and ',age.select), yaxt='n', ylab='Ratio')
axis(side=2, at=-c(-5, -0.5, 0, 0.5), label=round(exp(c(-5, -0.5, 0, 0.5)),2) )
}
par(mfrow=c(1,3))
lapply(ages, plot.rat)
```
Export for SATSCAN temporal clustering
```{r}
#write.csv(nyc, file='../Data/nyc_resp.csv')
```


```{r, echo=F, eval=F}
#Model with smoothing spline for year; cc=cyclic cubic spline
mod2<-gam(cbind(nyc$o65, nyc$u5)~  s(MMWRweek , bs='cc'), family=binomial, data=nyc)
plot(mod2)
pred2<-predict(mod2, type='response')
raw.ratio<-(nyc$`Ages 65+ years`+0.5)/(nyc$`Ages 0-4 years`+nyc$`Ages 65+ years`)
plot(nyc$date.index, pred2, type='p')
points(nyc$date.index,raw.ratio, col='red')
```

Disaggregated boroughs data. 
```{r}
bur1<- combo1.a[combo1.a$borough != 'Citywide',]
bur1$borough<-factor(bur1$borough)
bur1.fit<-bur1[bur1$date.index< 2020,]
```

Fit through Jan 2020, then extrapolate seasonal baseline forward
```{r}
mod2<-gam(cbind(bur1.fit$o65, bur1.fit$u5)~s(MMWRweek , bs='cc') +borough, family=binomial, data=bur1.fit)
pred2<-predict(mod2, newdata=bur1,type='response')
raw.ratio<-(bur1$`Ages 65+ years`+0.5)/(bur1$`Ages 0-4 years`+bur1$`Ages 65+ years`)
plot(bur1$date.index, pred2, type='p')
```

Pointwise log likelihood
```{r}
log.like.mod2<- -dbinom(bur1$o65,bur1$o65+ bur1$u5,pred2, log=T)
plot(bur1$date.index, log.like.mod2)
abline(v=2020, lty=2, col='gray', bty='l')
```


## Prepare data for JAGS

```{r}
ili<-combo1[combo1$syndrome==select.syndrome & combo1$age_group %in% c('Ages 0-4 years','Ages 65+ years'),  c('age_group','syndrome', 'borough','MMWRyear','MMWRweek','count')]
names(ili)<-c('age_group','syndrome', 'borough','MMWRyear','MMWRweek','ili')
ili$agec<-NA
ili$agec[ili$age_group=='Ages 0-4 years']<-1
ili$agec[ili$age_group=='Ages 65+ years']<-5
```

Use overall aggregate flu hospitalizations
```{r}
flu2<-flu[flu$AGE.CATEGORY %in% c('Overall'),]
#flu2$agec<-NA
#flu2$agec[flu2$AGE.CATEGORY=='0-4 yr']<-1
#flu2$agec[flu2$AGE.CATEGORY=='65+ yr']<-5
flu2<-flu2[,c('SEASON', 'MMWR.YEAR', 'MMWR.WEEK', 'WEEKLY.RATE')]
names(flu2)<-c('season', 'MMWRyear', 'MMWRweek','flu.rate')
flu2$flu.rate<-as.numeric(as.character(flu2$flu.rate))
flu2$flu.rate[is.na(flu2$flu.rate)]<-9999

```

```{r}
combo2<-merge(ili, flu2, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2<-merge(combo2, rsv, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2$flu.rate[is.na(combo2$flu.rate)]<-0
combo2$flu.rate[combo2$flu.rate==9999]<-NA #last few weeks
```


```{r}
combo2.sub<-combo2[, c('agec', 'MMWRyear', 'MMWRweek', 'borough', 'ili','flu.rate',"rsv.searches")]
ili.m<-melt(combo2.sub, id.vars=c('borough','agec','MMWRyear','MMWRweek'))
ili.a<-acast(ili.m, MMWRyear+MMWRweek ~borough~agec~variable , fun.aggregate = sum )
clean.array.jags<-ili.a[,c('Bronx','Brooklyn','Manhattan','Queens' ,'Staten Island'),c('1','5'),]
saveRDS(clean.array.jags,'./Data/jags.ds.rds')
```

```{r}
plot(scale(ili.a[,'Citywide','1','ili']), col='blue')
points(scale(ili.a[,'Citywide','1','flu.rate']), col='red')
points(scale(ili.a[,'Citywide','1','rsv.searches']), col='gray')

#Cross correlation
ccf(ili.a[,'Citywide','1','ili'], ili.a[,'Citywide','1',"rsv.searches"], na.action=na.omit)


```

