---
title: "covid_19_syndromic"
author: "Dan Weinberger"
date: "3/8/2020"
output: html_document
---

The goal of these analyses is to detect increases in syndromic disease activity associated with  COVID-19. As a test, we are using syndromic disease data from New York City. Data are stratified by day, borrough, age group. the data were downloaded from the [Epiquery website](https://a816-health.nyc.gov/hdi/epiquery/visualizations?PageType=ps&PopulationSource=Syndromic)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("..")) #sets Working directory to be 1 folder up

library(reshape2)
library(lubridate)
```

Import the data and remove the unnecesary rows. Note that the file downloads as a .csv format, but it is actually a tab-separated file.
```{r, eval=F}
ili1<- read.table('./Data/ILI_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 
resp1<-read.table('./Data/resp_2020_3_8.csv',sep="\t", header=TRUE, skipNul = T) 

combo1<-rbind.data.frame(ili1, resp1)

names(combo1)<-c('note','syndrome','geo.level','borough','ag.level','age_group','date','metric', 'count')

combo1<-combo1[,c('syndrome','borough','age_group','date','count')]
combo1$date<-as.Date(combo1$date, '%m/%d/%y')
combo1$count<-gsub(',', '', combo1$count) #remove commas from number
combo1$count<-as.numeric(as.character(combo1$count))
saveRDS(combo1,'./Data/ili_resp.rds')
```

Read in the data
```{r}
combo1<-readRDS('./Data/ili_resp.rds')
```

```{r}
#select.syndrome<-"Respiratory"
select.syndrome<-"Influenza-like illness (ILI)"

```

Reshape the data. For now, just focus on ILI
```{r}
combo1.m<-melt(combo1[combo1$syndrome==select.syndrome,], id.vars=c('syndrome', 'borough','age_group','date'))
combo1.a<-dcast(combo1.m, date +borough+syndrome~age_group )
```


```{r}
nyc<-combo1.a[combo1.a$borough=='Citywide',]
#nyc<-combo1.a[combo1.a$borough=='Manhattan',]
#nyc<-combo1.a[combo1.a$borough=='Queens',]
#nyc<-combo1.a[combo1.a$borough=='Bronx',]

```


```{r}
plot(nyc$date,nyc$"Ages 18-64 years", type='l', bty='l', main='ILI all ages in NYC')
```

```{r, fig.width=4, fig.height=6}
par(mfrow=c(2,1), mar=c(2,2,2,1))
plot(nyc$date,nyc$`Ages 0-4 years`, type='l', bty='l', main='ILI <5 years in NYC')
plot(nyc$date,nyc$`Ages 65+ years`, type='l', bty='l', main='ILI 65+ years in NYC')
```
Plot ratio of 65+ to younger age groups for ILI
```{r, fig.width=12, fig.height=4}
ages<-c("Ages 0-4 years", "Ages 5-17 years",  "Ages 18-64 years" )
plot.rat<-function(age.select){
nyc$age.rat<-(nyc$`Ages 65+ years`+0.5)/(nyc[,age.select]+0.5)
plot(nyc$date,log(nyc$age.rat), type='l', bty='l', main=paste0('ratio of 65+ and ',age.select), yaxt='n', ylab='Ratio')
axis(side=2, at=-c(-5, -0.5, 0, 0.5), label=round(exp(c(-5, -0.5, 0, 0.5)),2) )
}
par(mfrow=c(1,3))
lapply(ages, plot.rat)
```












