---
title: "ct_syndromic"
author: "Dan Weinberger"
date: "3/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = normalizePath("..")) #sets Working directory to be 1 folder up

library(reshape2)
library(lubridate)
library(INLA)
library(MMWRweek)
library(cdcfluview)
library(gtrendsR)
library(rjags)
library(HDInterval)
library(shiny)
library(pbapply)
library(zoo)
library(ciTools)
```

##Import and clean the data
```{r, eval=F}
ds1<-read.csv('../DO_NOT_SYNC/yale_view_2018_2019.csv')
zip.cw<-read.csv('../Data/zipcode-town.csv')

zip.cw$Zip.Code<-paste0('0',zip.cw$Zip.Code)
zip.cw[substr(zip.cw$Zip.Code,1,2) =='06',]
```

96% of entries have a CT ZIP code (06XXX)
```{r, eval=F}
ds1$classification<-toupper(ds1$classification)
ds1$ili<-grepl('ILI', ds1$classification)
ds1$cough<-grepl('COUGH', ds1$classification)
ds1$respiratory<-grepl('RESPIRATORY', ds1$classification)
ds1$circulatory<-grepl('CIRCULATORY', ds1$classification)
ds1$ili2<- grepl('FEVER/FLU', ds1$classification) 
datestr<-as.character(ds1$date_time_of_admission)
ds1$adate<-as.Date(substr(datestr,1,10))
ds1$agec<-NA
ds1$agec[ds1$age>=0  & ds1$age<5]<-1
ds1$agec[ds1$age>=5  & ds1$age<18]<-2
ds1$agec[ds1$age>=18 & ds1$age<39]<-3
ds1$agec[ds1$age>=40 & ds1$age<64]<-4
ds1$agec[ds1$age>=65 & ds1$age<110]<-5
ds1$zipcode<-as.character(ds1$zipcode)
ds1$all.visits<-1
ds2<-merge(ds1, zip.cw, by.x='zipcode', by.y='Zip.Code')

#last.date<-ds2[ds2$adate==as.Date('2020-03-09') & ds2$agec==5 & ds2$County=='Fairfield' ,c('classification'), drop=F]
```

```{r, eval=F}
ds1.m<-melt(ds2[,c('adate', 'County','agec','ili','cough','respiratory','circulatory','ili2','all.visits')], id.vars=c('adate','County','agec'))
last.date<- max(ds1.m$adate)
ds1.m<-ds1.m[ds1.m$adate != last.date,] #remove last day from the dataset,assuming it is incomplete
saveRDS(ds1.m, '../DO_NOT_SYNC/county_ts_ct.rds')
ds1.c<-acast(ds1.m, agec~adate ~County~variable, fun.aggregate = sum)
```

```{r}
ds1.m<-readRDS('../DO_NOT_SYNC/county_ts_ct.rds')
ds1.c<-acast(ds1.m, agec~adate ~County~variable, fun.aggregate = sum)
saveRDS(ds1.c,'./private_data/shiny_data.rds') #save data for the shiny app

```

## View data
```{r}
counties<-dimnames(ds1.c)[[3]]
ages<-dimnames(ds1.c)[[1]]
#syndromes<- dimnames(ds1.c)[[4]]
syndromes<-'ili2'
for(j in syndromes){
for(i in counties){
par(mfrow=c(2,3))  
for(k in ages){
  plot(ds1.c[k,,i,j]/ds1.c[k,,i,'all.visits'], type='l', bty='l', main=paste(j,i,k))
  }
 }
}


```




##Import auxillary data
Pulls in google searches for RSV in CT can get 5 years of historical data at weekly reslution (switches to monthly for >5 years)
```{r}
rsv.down<-gtrends(keyword = 'RSV', geo="US-CT", time = "today+5-y",
gprop = c("web") , category = 0,  low_search_volume = FALSE)
saveRDS(rsv.down,'../Data/rsv_searches_ct.rds')
rsv<-rsv.down$interest_over_time[,c('date','hits')]
rsv$date<-as.Date(rsv$date)
mmwr.week.rsv<-MMWRweek(rsv$date)[,c('MMWRyear','MMWRweek')]
rsv<-cbind.data.frame(rsv,mmwr.week.rsv)
names(rsv)<-c('date','rsv.searches','MMWRyear','MMWRweek')
```

flusurvnet data
https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html

```{r}
flu<-read.csv('../Data/FluSurveillance_EIP_Connecticut_Data.csv', skip=2)
flu2<-flu[flu$AGE.CATEGORY %in% c('Overall'),]
#flu2$agec<-NA
#flu2$agec[flu2$AGE.CATEGORY=='0-4 yr']<-1
#flu2$agec[flu2$AGE.CATEGORY=='65+ yr']<-5
flu2<-flu2[,c('SEASON', 'MMWR.YEAR', 'MMWR.WEEK', 'WEEKLY.RATE')]
names(flu2)<-c('season', 'MMWRyear', 'MMWRweek','flu.rate')
flu2$flu.rate<-as.numeric(as.character(flu2$flu.rate))
flu2$flu.rate[is.na(flu2$flu.rate)]<-9999
```

Aggregate by week and merge auxillary variables
```{r}
ds1.m.wk<-ds1.m
#ds1.m.wk$adate<-floor_date(ds1.m.wk$adate, unit='week')
ds1.m.wk$adate<-ds1.m.wk$adate
ds1.df<-dcast(ds1.m.wk, agec+adate +County~variable, fun.aggregate = sum)
mmwr.date<-MMWRweek(ds1.df$adate)
ds1.df<-cbind.data.frame(ds1.df,mmwr.date)
combo2<-merge(ds1.df, flu2, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2<-merge(combo2, rsv, by=c('MMWRyear','MMWRweek'), all.x=T)
combo2$flu.rate[is.na(combo2$flu.rate)]<-0
combo2$flu.rate[combo2$flu.rate==9999]<-NA #last few weeks

combo2.sub<-combo2[, c('agec', 'adate','MMWRyear', 'MMWRweek', 'County', 'ili2','cough','respiratory','circulatory','flu.rate',"rsv.searches",'all.visits')]
ili.m<-melt(combo2.sub, id.vars=c('County','agec','adate','MMWRyear','MMWRweek'))
ili.a<-acast(ili.m, adate+MMWRyear+MMWRweek ~County~agec~variable , fun.aggregate = sum )
dimnames(ili.a)[[1]]<-substr(dimnames(ili.a)[[1]],1,10)
```

setup epiyear index
```{r}
date.string<-as.Date(dimnames(ili.a)[[1]])
month<-month(date.string)
epiyr<-year(date.string)
epiyr[month<=6] <- epiyr[month<=6]-1
epiyr.index<-epiyr-min(epiyr)+1
weekN<-MMWRweek(date.string)[,'MMWRweek']
day.of.year<-yday(date.string)
```

## Evaluate results after controlling for flu and RSV
```{r}
counties.to.test<-dimnames(ili.a)[[2]]

gam.func<-function(x.test, age.test, syndrome, all.cause.name){
clean.array.citywide<-ili.a[,x.test,,]
      epiyr.index.f<-as.factor(epiyr.index)
      epiyr.index.f2<-as.factor(epiyr.index)
      
      y.age = t(clean.array.citywide[,age.test,syndrome])
      offset.age = t(clean.array.citywide[,age.test,all.cause.name])

      y.age.fit<-y.age[1,]
      y.age.fit[epiyr.index==2019 & day.of.year>=60] <- NA #extrapolate after day 60
     
            sqrt.rsv =sqrt(clean.array.citywide[,age.test,'rsv.searches']) #same for all ages and boroughs
            sqrt.flu =sqrt(clean.array.citywide[,age.test,'flu.rate']) #same for all ages and boroughs
            
            
            sqrt.flu<- na.locf(sqrt.flu)  #fill in missing observations for flu at end of TS with most recent observed values
      t2<-1:length(y.age)
      # 
      # sin1<-sin(2*pi*t2/365.25)
      # cos1<-cos(2*pi*t2/365.25)
      # sin2<-sin(2*pi*t2*2/365.25)
      # cos2<-cos(2*pi*t2*2/365.25)
      # sin3<-sin(2*pi*t2*3/365.25)
      # cos3<-cos(2*pi*t2*3/365.25)
      # 

      
      ds.glm<-cbind.data.frame('day.of.year'=day.of.year,'y.age'=y.age[1,],sqrt.rsv, sqrt.flu,  t2,epiyr.index.f)
      # form1<-as.formula(paste0('y.age',"~ epiyr.index.f*sqrt.rsv +   #rsv effect varies by epiyr
      #              epiyr.index.f*sqrt.flu + #flu effect, varies by epiyear
      #              t2+                      #linear trend
      #              s(day.of.year, bs='cc', k=6)" #seasonality
      #              ))
      #  mod1<-gam(form1,
      #         family='poisson', data=ds.glm,offset=log(offset.age[1,]+0.5))
      # pred1<-predict(mod1, type='response', newdata = ds.glm)*(offset.age[1,]+0.5)
      form1<-as.formula(paste0('y.age',"~ epiyr.index.f*sqrt.rsv +   #rsv effect varies by epiyr
                   epiyr.index.f*sqrt.flu + #flu effect, varies by epiyear
                   #t2+                      #linear trend
                   sin1+cos1 +sin2+cos2+sin3+cos2 +log(offset.age[1,]+0.5) "
                   ))
        mod1<-glm(form1,
                    family='poisson', data=ds.glm)
      pred1<-predict(mod1, type='response', newdata = ds.glm)

     my_data_with_pi <- add_pi(ds.glm, mod1, names = c("lpb", "upb"))

      resid1<-log((y.age[1,]+0.5)/(pred1+0.5))
      
      out.list<-list(y=y.age[1,], pred=my_data_with_pi$pred, resid1=resid1, upi=my_data_with_pi$upb, lpi=my_data_with_pi$lpb)
      return(out.list)
}


syndromes<- c('ili2', 'respiratory','cough') 
ages <-   c('1','2','3','4','5','NA')

all.gam.res<- pblapply(syndromes, function(x){
  ww<- lapply(ages, function(y){
    lapply(counties.to.test, gam.func, age.test=y, syndrome=x, all.cause.name='all.visits')
  }
  ) 
  names(ww)<- ages
  return(ww)
  }
)
names(all.gam.res)<-syndromes

```


Next extract residuals and fitted values
```{r, fig.width=8, fig.height=6}

#Residuals for all age groups for ILI
ili2.resid<- sapply(all.gam.res[['ili2']], function(x) sapply(x,'[[','resid1'), simplify='array')
dimnames(ili2.resid)[[2]]<-counties.to.test
ili2.pred<- sapply(all.gam.res[['ili2']], function(x) sapply(x,'[[','pred'), simplify='array')
dimnames(ili2.pred)[[2]]<-counties.to.test

ili2.pred.lcl<- sapply(all.gam.res[['ili2']], function(x) sapply(x,'[[','lpi'), simplify='array')
dimnames(ili2.pred.lcl)[[2]]<-counties.to.test

ili2.pred.ucl<- sapply(all.gam.res[['ili2']], function(x) sapply(x,'[[','upi'), simplify='array')
dimnames(ili2.pred.ucl)[[2]]<-counties.to.test

obs<- sapply(all.gam.res[['ili2']], function(x) sapply(x,'[[','y'), simplify='array')
dimnames(obs)[[2]]<-counties.to.test
dates<-as.Date(dimnames(ili.a)[[1]])

par(mfrow=c(2,3), mar=c(3,2,1,1))
for(i in 1:5){
  matplot(dates,ili2.resid[,1,i], type='l', bty='l', ylab='Residual')
  abline(h=1)
}

par(mfrow=c(2,3), mar=c(3,2,1,1))
for(i in 1:5){
  matplot(dates[dates>=as.Date('2020-01-01')],obs[dates>=as.Date('2020-01-01'),,i], pch=16, bty='l', ylab='Fitted')
  matplot(dates[dates>=as.Date('2020-01-01')],ili2.pred[dates>=as.Date('2020-01-01'),,i], type='l', bty='l', ylab='Fitted', add=T)
}

pdf(file='./private_data/ct_county_age.pdf')
for(j in counties.to.test){
par(mfrow=c(2,3), mar=c(3,2,1,1))
for(i in 1:5){
  plot(dates[dates>=as.Date('2020-01-01')],obs[dates>=as.Date('2020-01-01'),j,i], type='l', bty='l', ylab='Fitted', main=paste(j, age.labels[i]), ylim=range(c(ili2.pred.lcl[,j,i],ili2.pred.ucl[,j,i],ili2.pred[,j,i])))
  points(dates[dates>=as.Date('2020-01-01')],ili2.pred[dates>=as.Date('2020-01-01'),j,i], type='l', col='red', lty=3 )
    points(dates[dates>=as.Date('2020-01-01')],ili2.pred.lcl[dates>=as.Date('2020-01-01'),j,i], type='l', col='red', lty=3 )
    points(dates[dates>=as.Date('2020-01-01')],ili2.pred.ucl[dates>=as.Date('2020-01-01'),j,i], type='l', col='red', lty=3 )
  abline(v=790)
 }
}
dev.off()

```
Age difference in residuals 40-64 vs 65+
```{r, fig.width=12, fig.height=8}
diff.resid<-ili2.resid[,,4] - ili2.resid[,,1]
par(mfrow=c(2,4))
lapply(counties.to.test, function(x) plot(diff.resid[,x], type='l', bty='l', main=x))
```





##Temporal SATSCAN by county and syndrome
Do a Poisson temporal satscan, adjusting for fitted values as the population file
```{r}

syndromes<- c('ili2', 'respiratory','cough') 
ages <-   c('1','2','3','4','5')
counties.to.test<-dimnames(ili.a)[[2]]

satscan.fun<-function(syndrome, county.select,age.select){

obs<- sapply(all.gam.res[[syndrome]], function(x) sapply(x,'[[','y'), simplify='array')
dimnames(obs)[[2]]<-counties.to.test

case.pred<- sapply(all.gam.res[[syndrome]], function(x) sapply(x,'[[','pred'), simplify='array')
dimnames(case.pred)[[2]]<-counties.to.test

mycas = data.frame(locid=1:dim(case.pred)[[1]] ,basecas=obs[,county.select,age.select], day=as.Date(dimnames(case.pred)[[1]]))
mypop<- data.frame(locid=1:dim(case.pred)[[1]] ,day=as.Date(dimnames(case.pred)[[1]]),basepop=case.pred[,county.select,age.select] )
mygeo<- data.frame(locid=1:dim(case.pred)[[1]] ,lat=1:dim(case.pred)[[1]], long=1:dim(case.pred)[[1]])

#Write satscan files

td = tempdir()
write.cas(mycas, location = td, file = "mycas")
write.pop(mypop, location = td, file = "mypop")
write.geo(mygeo, location = td, file = "mygeo")

#write.geo(mygeo, location = '../satscan/data/', file = "geo.spacetime.ct", userownames=F)
#write.cas(mycas, location = '../satscan/data/', file = "ili.temporal")
#write.pop(mypop, location = '../satscan/data/', file = "ili.expected")

#Parameter files

first.date<-as.character(min(mycas$day))
first.date<-gsub('-', '/',first.date)
last.date<-as.character(max(mycas$day))
last.date<-gsub('-', '/',last.date)
invisible(ss.options(reset=TRUE))
ss.options(list(CaseFile="mycas.cas",PopulationFile='mypop.pop', CoordinatesFile="mygeo.geo", PrecisionCaseTimes=3))
ss.options(list(StartDate=first.date, CoordinatesType=1, TimeAggregationUnits=3,TimeAggregationLength=5))
ss.options(list(EndDate=last.date,  AnalysisType=6, ModelType=0,ScanAreas=1)) 
ss.options(list(ProspectiveStartDate="2019/01/01", ReportGiniClusters="n", LogRunToHistoryFile="n"))

write.ss.prm(td, "mybase")
#write.ss.prm('../satscan/satscan parameters', "temporal adjust rsatscan")

mybase = satscan(td, "mybase")
return(mybase)
}

all.satscan.res<- pblapply(syndromes, function(x){
  ww<- lapply(ages, function(y){
    lapply(counties.to.test, satscan.fun, age.select=y, syndrome=x)
  }
  ) 
  names(ww)<- ages
  return(ww)
  }
)
names(all.satscan.res)<-syndromes

#Extract results
ili.satscan<- sapply(all.satscan.res[['ili2']], function(x) sapply(x,'[[','col', simplify=F), simplify='F')
all.ages.ili<-lapply(all.satscan.res[['ili2']], function(y) sapply(y, function(x) apply(x$col,2,function(z) as.character(z))))


```


## Do a Normal SATSCAN for difference in residuals
```{r}

```

